### 2016-07-06 related-work.tex

\section{Related Work}
\label{sec:relatedwork}

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for formal
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace; their framework also requires application invariants be
expressed in terms of the visibility relations used to define
isolation and consistency properties.  Cerone \emph{et
  al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
visibility, but their specifications are also for complete traces, and
assumes all transactions execute using either {\sc si} or {\sc ser},
which is often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types, none of them
are equipped with a reasoning framework that can use such
specifications to verify programs under weak isolation.

\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
\emph{et al.} propose a theory to characterize non-serializable
executions that arise under {\sc si}. Fekete~\cite{fekete2005} also
proposes an algorithm that allocates either {\sc si} or {\sc ser}
isolation levels to transactions while guaranteeing the
serializability. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve
on Adya's {\sc si} specification and use it to derive a static
analysis that determines the safety of substituting {\sc si} with a
weaker variant called \iso{Parallel Snapshot Isolation}~\cite{psi}.
These efforts focus on establishing the equivalence of executions
between a pair of isolation levels, without taking application
invariants into account.  Bernstein \emph{et al.}~\cite{bern2000}
propose informal semantic conditions to ensure the satisfaction of
application invariants under weaker isolation levels.  All these
techniques are tailor-made for a finite set of well-understood
isolation levels (rooted in~\cite{berenson}) under a pre-defined store
consistency model (usually {\sc sc}).

\paragraph{Reasoning under weak consistency} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{lbc16,gotsmanpopl16}). The key technical issue in
these efforts involves proving that atomic and fully isolated
operations preserve application invariants when executed on these
consistency levels.  In contrast, we admit weakly isolated
transactions, and our system model accepts \emph{specifications} of
consistency and isolation levels drawn from an expressive logic.
Gotsman \emph{et al.}~\cite{gotsmanpopl16} adapts \iso{Parallel
Snapshot Isolation} to the aforementioned setting by interpreting it
as a consistency level that serializes writes to objects; a dedicated
proof rule is developed to help prove prove program invariants hold
under this model. By parameterizing our proof system over a gamut of
weak isolation specifications, we avoid the need to define a separate
proof rule for each new isolation level we may encounter.

\paragraph{Reasoning under relaxed memory} The reasoning mechanisms
used to describe and prove properties about weakly-isolated
transactions bear resemblance to those used to formalize relaxed
memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
\emph{et al.}~\cite{ferreira10} propose a parameterized operational
semantics for relaxed memory models, but the parameterization is over
a relation between relaxed memory programs and related {\sc sc}
programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
  memory model} for Java that defines an axiomatic definition for the
JMM in terms of memory reorderings, and an operational instantiation
consistent with the TSO memory model, and demonstrate the equivalence
of the two definitions.

>>>>

\vspace*{-8pt}
\section{Related Work}
\label{sec:relatedwork}

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, and the kinds of
dependencies that are forbidden in each case. The operational nature
of Adya's specifications make them suitable for runtime monitoring and
anomaly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
the declarative nature of our specifications make them suitable for
formal reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, making it difficult
to reason about a program as it builds the trace; their framework also
requires application invariants be expressed in terms of low-level
visibility relations.  Cerone \emph{et al.}~\cite{gotsmanconcur15} specify
isolation levels with atomic visibility, but their specifications are
also for complete traces, and assumes all transactions execute using
either {\sc si} or {\sc ser}, which is often not the case in
practice. While all the aforementioned specification frameworks use
the vocabulary introduced in~\cite{burckhardt14} to specify replicated
data types, none of them are equipped with a reasoning framework that
can use such specifications to verify programs under weak isolation.

\vspace*{-4pt}
\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
\emph{et al.} propose a theory to characterize non-serializable
executions that arise under {\sc si}. Fekete~\cite{fekete2005} also
proposes an algorithm that allocates either {\sc si} or {\sc ser}
isolation levels to transactions while guaranteeing
serializability. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve
on Adya's {\sc si} specification and use it to derive a static
analysis that determines the safety of substituting {\sc si} with a
weaker variant called \iso{Parallel Snapshot Isolation}~\cite{psi}.
These efforts focus on establishing the equivalence of executions
between a pair of isolation levels, without taking application
invariants into account.  Bernstein \emph{et al.}~\cite{bern2000}
propose informal semantic conditions to ensure the satisfaction of
application invariants under weaker isolation levels.  All these
techniques are tailor-made for a finite set of well-understood
isolation levels (rooted in~\cite{berenson}) under a pre-defined store
consistency model.

\vspace*{-4pt}
\paragraph{Reasoning under weak consistency} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
  redblueosdi, ecinec}. All of them assume a system model that offers
a choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
  ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{lbc16,gotsmanpopl16}). All these efforts involve
proving that atomic and fully isolated operations preserve application
invariants when executed under these consistency levels.  In contrast,
we admit weakly-isolated transactions, and our system model accepts
\emph{specifications} of consistency and isolation levels drawn from
an expressive logic.  Gotsman \emph{et al.}~\cite{gotsmanpopl16}
adapts \iso{Parallel Snapshot Isolation} to the aforementioned setting
by interpreting it as a consistency level that serializes writes to
objects; a dedicated proof rule is developed to help prove prove
program invariants hold under this model. By parameterizing our proof
system over a gamut of weak isolation specifications, we avoid the
need to define a separate proof rule for each new isolation level we
may encounter.

\vspace*{-4pt}
\paragraph{Reasoning under relaxed memory} The reasoning mechanisms
used to describe and prove properties about weakly-isolated
transactions bear some resemblance to those used to formalize relaxed
memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
\emph{et al.}~\cite{ferreira10} propose a parameterized operational
semantics for relaxed memory models, but the parameterization is over
a relation between relaxed memory programs and related {\sc sc}
programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
  memory model} for Java that defines an axiomatic definition for the
JMM in terms of memory reorderings, and an operational instantiation
consistent with the TSO memory model.

### 2016-07-06 store-consistency.tex

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximul visibility weakening of {\sc sc} w.r.t. $\I_c$, as
described in \S\ref{sec:store-consistency}. In concrete terms, this
means that an effect added to $\E.\A$ by the current thread or a
concurrent thread immediately becomes visible to the subsequent
operations unless we produce a witness that such visibility violates
$\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note that the language of
our specifications is a decidable subset of the first-order logic. In
particular, our specifications and rely-guarantee rules are carefully
crafted to avoid generating proof obligations with transitive closures
and quantification over infinite domains. The domain of assertions is
often linear arithmetic, which is also decidable. Consequently, the
generated proof obligations can often be discharged via an SMT solver.
We nonetheless give full details in the following.

>>>>

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximum visibility weakening of {\sc sc} w.r.t. $\I_c$, as
described in \S\ref{sec:store-consistency}. In concrete terms, this
means that an effect added to $\E.\A$ by the current thread or a
concurrent thread immediately becomes visible to the subsequent
operations unless we produce a witness that such visibility violates
$\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note that the language of
our specifications is a decidable subset of the first-order logic. In
particular, our specifications and rely-guarantee rules are carefully
crafted to avoid generating proof obligations with transitive closures
and quantification over infinite domains. The domain of assertions is
often linear arithmetic, which is also decidable. Consequently, the
generated proof obligations can often be discharged via an SMT solver.
